Determining the existance and direction of causality in the face of measurement error

### 2016-05-09 docs/manuscript.tex
finished new draft

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Determining the existance and direction of causality in the face of measurement error},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existance and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{05 May 2016}

@@

With our ability to characterize the human phenome ever improving it is
becoming increasingly common to use statistical tests to infer the
causal relationships between correlated variables. A simple method for
inferring if an exposure is on the causal pathway to an outcome is
through mediation analysis, where the effect of an instrument on the
outcome is tested before and after adjusting for the exposure. We show
that in the face of measurement error this can lead to erroneous
results, and that increasing sample size, rather than reducing bias, can
often have the alarming effect of increasing certainty in the wrong
answer. Commonly used tools that are predicated on this methodology such
as the causal inference test (CIT) are shown to be susceptible. We
demonstrate that Mendelian randomization (MR) is a method for causal
inference that is robust to measurement error, and that a simple
extension to MR can provide a method to estimate the direction of
causality between correlated variables even when the biology of the
instrument is unknown. We argue that because measurement error is
ubiquitous in phenotypic data, mediation-based causal inference should
be treated with caution.

@@

With the explosion in data that assess the human phenome, we are
presented with the challenge of understanding the causal nature of the
emerging putitive correlations between measures of interest. Mediation
is a commonly used statistical method for making causal inference in
observational data\textsuperscript{1--4}, and it exists in many forms,
from simple regression based systems to structural equation modeling
(SEM)\textsuperscript{5}. The principle behind mediation analysis can be
explained as follows. Supposing an exposure has a causal effect on an
outcome, then an `instrument' that causes the exposure (a single
nucleotide polymorophism (SNP), for example) should also influence the
outcome. Therefore the influence of the instrument on the outcome
conditional on the exposure should be zero. This forms the basis of a
number of methods, such as the regression-based causal inference test
(CIT)\textsuperscript{1}, an SEM implementation in the NEO
software\textsuperscript{5}, and various Bayesian methods, which have
been employed by a number of recent publications that make causal
inferences, often in large scale `omics datasets\textsuperscript{6--8}.

While mediation based approaches are simple and convenient to implement,
it is important to note where they may lead to unreliable results. One
such mechanism is measurement error\textsuperscript{9--12}, which has
been noted to be an explicit issue with CIT style
approaches\textsuperscript{13,14}. Measurement (or observational) error
is the difference between a measured value of a quantity and its true
value. Such variability can arise through a whole plethora of
mechanisms, which are often unique to the study design and difficult to
avoid\textsuperscript{15,16}. Array technology is now commonly used to
obtain high throughput phenotyping at low cost, but they come with the
problem of having imperfect sensitivity, so for example methylation
levels as measured by the Illumina450k chip are prone to have some
amount of noise around the true value\textsuperscript{17,18}.
Sensitivity is also an issue, for example if the unit of measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{19}.

Measurement error will of course arise in more low-tech data too, for
example when measuring body mass index (BMI) one is typically interested
in using this as a proxy for obesity, but it is clear that the
correlation between BMI and underlying adiposity is not
perfect\textsuperscript{20}. A similar problem of biological
misspecification is unavoidable in disease diagnosis. Measurement error
can also be introduced after the data has been collected, for example
the transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically have both bias
and imprecision compared to the original variable. The sources of
measurement error are not limited to this list\textsuperscript{16}, and
its impact has been explored in the context of mediation analysis in the
epidemiological literature extensively\textsuperscript{9,12}.

An alternative statistical method that gained prominence as a solution
to the problem of measurement error is Mendelian randomisation
(MR)\textsuperscript{21,22}. Here an instrumental variable (typically a
SNP) that has a robost, causal association with the exposure is used to
stand for the exposure in an association test with the outcome. If there
is no measurement error in the SNP then the causal association between
the exposure and the outcome can be estimated by scaling the association
between the SNP and the outcome by the association between the SNP and
the exposure. Indeed this approach was in part pioneered (in the form of
instrumental variable (IV) analysis) as a correction for measurement
error\textsuperscript{23}, but it also has the useful property that it
guards against unmeasured confounders driving the association between
exposure and outcome; and if the biological effect of the SNP is
understood then it can also aid against reverse causality driving the
association.

Here, however, arises a potential issue: often the biological effect of
a SNP is not known. If there exists a putative association between two
variables, with a SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e., for which of the two variables is the
SNP a valid instrument?). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. For example genome wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{24}. To illustrate a trivial example
- the same FTO variant that is famously associated with BMI is also
associated with C-reactive protein (CRP). It is known that this is
because BMI causes CRP\textsuperscript{25}, but supposing we only had
\emph{a priori} knowledge that FTO associated with CRP, and we used it
to instrument CRP in an MR analysis of CRP against BMI, we would infer
that CRP causes BMI. Another situation is where the causal direction
between 'omic measures need to be determined, for example if a DNA
methylation probe is associated with expression of an adjacent gene,
then is a cis-acting SNP an instrument for the DNA methylation level, or
the gene expression level\textsuperscript{8}?

Mediation based methods, for example CIT and other related approaches
typically attempt to use the data to resolve this issue using the
following logic: if the SNP-exposure association is stronger than the
SNP-outcome association then the causal effect must be in the direction
of exposure to outcome. The same logic can indeed be applied to MR, but
often when applied the sensitivity of these approaches to measurement
error is not considered.

When being objective one must assume that measurement error is
ubiquitous, and any measured variable is only an imperfect proxy of the
biological quantity that the researcher intended to obtain. Here we show
using theory and simulations how measurement error can lead to
unreliable causal inference in the mediation-based CIT method, and we
propose an extension to MR that, although does not eliminate the problem
of measurement error, provides a framework in which to ascertain the
causal direction when the biology of the instrument is not fully
understood.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{26} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{27}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\) by making inferences based on a
number of conditions that we examine in the results. To infer the
direction of causality using the CIT method, an omnibus p-value
generated by CIT, \(p_{CIT}\), was estimated for the correct model of
\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
some significance threshold \(\alpha\), causal direction was inferred
based on the following scenarios:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then the correct model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then the incorrect model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then the an alternative causal model is accepted
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then no inference is made
\end{enumerate}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value.

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population\textsuperscript{28}. It
is calculated as

@@

and a p-value is generated from the \(Z\) value to indicate confidence
of difference in correlations \(\rho_{gx}\) and \(\rho_{gy}\). The test
for a causal association is then obtained by performing two sample MR in
the direction that is predicted to be the correct model based on the Z
score test. For the purposes of compiling simulation results, the test
is deemed to have obtained a robust association if the Z score test has
a p-value \(< 0.05\) and the two sample MR analysis has a p-value
\(< 0.05\), though in practice we do not advocate using an arbitrary
\(p\) value threshold for denoting significance in making causal
inference.

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{29} setting, where the Steiger test of two
independent correlations is applied instead of the case detailed here
where correlations share one variable in common. An advantage of using
the Steiger test in the two sample context is that it can compare
correlations in independent samples where sample sizes are different.

@@

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \beta_g g + \epsilon_g \\
x_O & = \beta_{mx} x + \epsilon_{mx} \\
y & = \beta_x x + \epsilon_x \\
y_O & = \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

All \(\beta\) values were set to 1. Values of \(\epsilon_*\) were
generated such that

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
var(\epsilon_{mx}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
var(\epsilon_{my}) & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Each of these sets of
variables was performed 100 times, and the CIT and MR methods were
applied to each in order to evaluate the causal association of the
simualted variables. Similar patterns of results were obtained for
different values of \(cor(g, x)\).

@@

\subsubsection{The influence of measurement error on
CIT}\label{the-influence-of-measurement-error-on-cit}

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_O=f(x)\). For
example, we can define \(f(x) = \alpha_m + \beta_m x + \epsilon_m\),
where \(\alpha_m\) and \(\beta_m\) influence the bias in the measurement
of \(x\), and \(\epsilon_m\) represents the imprecision in the
measurement of \(x\). Here the true value of the exposure is partially
explained by the genetic instrument, \(g\), such that

@@

where \(\beta_g\) is the effect of the SNP on the exposure, and
\(\epsilon_g\) is the residual value of \(x\); and the outcome is
partially explained by the exposure

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(g\) is associated with \(x|y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition is necessary for causal
inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement error
is introduced we can show using basic covariance properties (Appendix 1)
that

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Determining the existance and direction of causality in the face of measurement error},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{Determining the existance and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{09 May 2016}

@@

With our ability to characterize the human phenome ever improving it is
becoming increasingly common to use statistical tests to infer the
causal relationships between correlated variables. A simple and commonly
used method for inferring if an exposure is on the causal pathway to an
outcome is through using genetic instruments in mediation analysis,
where the effect of an instrument on the outcome is tested before and
after adjusting for the exposure. We show that in the face of
measurement error this can lead to erroneous results, and that
increasing sample size, rather than reducing bias, can often have the
effect of increasing certainty in the wrong answer. Commonly used tools
that are predicated on this methodology such as the causal inference
test (CIT) are shown to be susceptible. We demonstrate that Mendelian
randomization (MR) is a method for causal inference that is robust to
measurement error, and that a simple extension to MR can provide a
method to estimate the direction of causality between correlated
variables even when the biology of the instrument is unknown. Using this
method we demonstrate that for correlations between DNA methylation and
gene expression levels, DNA methylation is likely to be the causal
factor. We show that using MR to infer causal directions is
straightforward, but because measurement error is ubiquitous in
observational data, mediation-based causal inference in particular
should be treated with caution.

@@

The last decade has seem a rapid acceleration in technological advances
that enable us to measure the human phenome in ever increasing detail,
furnishing biologists with the tantalising opportunity to map the
molecular basis of complex traits. This does, however, require
overcoming the statistical challenge of correctly defining the causal
architectures that underlie these high dimensional data. For example,
identifying metabolic perturbations that lead to
disease\textsuperscript{1}, or understanding the influence of epigenomic
influences on gene expression\textsuperscript{2} all depend on
understanding the causal nature of any putative observational
associations. Despite the ubiquity of the question, commonly used
observational study data are notoriously susceptible to pitfalls in
making causal inference, with simple regression-based techniques unable
to distinguish a true causal association from reverse causation or
confounding\textsuperscript{3}.

An established statistical design that can resolve this problem is
randomisation. Given the hypothesis that an exposure is causally related
to an outcome, randomisation can be employed to assess the causal nature
of the association by randomly splitting the sample into two groups,
subjecting one group to the exposure and leaving the other as a control.
The association between the exposure and the outcome in this setting
provides a robust estimate of the causal relationship. This is the
theoretical basis behind randomised control trials, but in practice
randomisation is often impossible to implement in an experimental
context due to cost, scale or inability to manipulate the exposure. The
principle, however, can be employed in extant observational data through
the use of genetic variants associated with the exposure (instruments),
where the inheritance of an allele serves as a random lifetime
allocation of differential exposure levels\textsuperscript{3}. Two
statistical approaches to exploiting the properties of genetic
instruments are widely used: mediation and Mendelian randomisation (MR).

Mediation is a commonly used statistical method for making causal
inference in observational data, and it exists in many forms, from
regression based systems\textsuperscript{4} to structural equation
modeling (SEM)\textsuperscript{5}. In its simplest form mediation
analysis seeks to explain the mechanism that underlies an observed
relationship between an independent and a dependent variable by
introducing a third, `mediator' variable. If the association between the
independent and dependent variables disappears when conditioning on the
mediator variable then this provides evidence that the independent
variable influences the dependent variable through the mediator.

This simple form, however, is prone to being unreliable. Alternative
models of reverse causation or confounding are often indistinguishable
from the hypothesised model of the independent variable causing the
dependent variable via the mediator\textsuperscript{6}. Introducing
pseudo-randomisation by using a genetic instrument as the independent
variable can be used to improve on these issues. Supposing an exposure
has a causal effect on an outcome, then an `instrument' that causes the
exposure (a single nucleotide polymorophism (SNP), for example) should
also influence the outcome. Therefore the influence of the instrument on
the outcome conditional on the exposure should be zero. This forms the
basis of a number of methods such as genetical
genomics\textsuperscript{7}, the regression-based causal inference test
(CIT)\textsuperscript{8,9}, an SEM implementation in the NEO
software\textsuperscript{5}, and various other methods including
Bayesian approaches\textsuperscript{10}. They have been employed by a
number of recent publications that make causal inferences, often in
large scale `omics datasets\textsuperscript{10--15}.

MR employs the same data under but uses a different approach in how the
instrumental variable is employed. Here an instrumental variable
(typically a SNP) that has a robost, direct association with the
exposure is used as a surrogate for the exposure in an association test
with the outcome, avoiding the problems of confounding or reverse
causation by testing for association between the outcome and the SNP.
The causal association between the exposure and the outcome can be
estimated by scaling the association between the SNP and the outcome by
the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation and MR
models have properties that confer some advantages and some
disadvantages. The main advantage of mediation over MR is that the true
underlying biology of the genetic instrument doesn't necessarily need to
be known as long as it associates with the putative exposure. In the CIT
framework (described fully in the Methods) for example, the test
statistic is different if you test for the exposure causing the outcome
or the outcome causing the exposure, allowing the researcher to infer
the direction of causality between two variables by performing the test
in both directions and choosing the model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of
causality the instrument must be known to influence the exposure
directly, associating with the outcome only through the exposure. This
requirement can be problematic because if there exists a putative
association between two variables, with the instrumental SNP being
robustly associated with each, it can be difficult to determine which of
the two variables is subject to the direct effect of the SNP (i.e., for
which of the two variables is the SNP a valid instrument?). By
definition, we expect that if the association is causal then a SNP for
the exposure will be associated with the outcome, so if the researcher
erroneously uses the SNP as an instrument for the outcome then they are
likely to see an apparently robust causal association of outcome on
exposure. Such a situation can arise in many scenarios. Genome wide
association studies (GWASs) that identify genetic associations for
complex traits are, by design, hypothesis free and agnostic of genomic
function, and it often takes years of follow up studies to understand
the biological nature of a putative GWAS hit\textsuperscript{16}. For
example, the same FTO variant that is associated with body mass index
(BMI) is also associated with C-reactive protein (CRP). It is known that
this is because BMI causes CRP\textsuperscript{17}, but supposing we
only had \emph{a priori} knowledge that FTO associated with CRP, and we
used it to instrument CRP in an MR analysis of CRP against BMI, we would
mistakenly infer that CRP causes BMI. Another situation is where the
causal directions between 'omic measures need to be determined, for
example if a DNA methylation probe is associated with expression of an
adjacent gene, then is a cis-acting SNP an instrument for the DNA
methylation level, or the gene expression level\textsuperscript{18}?

MR does however have some important advantages over mediation based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, using
mediation requires that the exposure, outcome and instrument variables
are all measured in the same data, whereas recent extensions to MR
circumvent this requirement, allowing causal inference to be drawn when
exposure variables and outcome variables are measured in different
samples\textsuperscript{19}. This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes.
Third, and though this is often underappreciated in fields outside
classical epidemiology it is of crucial importance, MR is substantially
more robust to there being measurement error in the exposure and outcome
variables. Indeed instrumental variable (IV) analysis was in part
initially introduced as a correction for measurement
error\textsuperscript{20}, whereas it has been noted that both basic
mediation\textsuperscript{21--24} and mediation methods that use
instrumental variables\textsuperscript{25,26} are prone to be unreliable
in its presence. We argue that this is a serious limitation.

Measurement (or observational) error is the difference between the
measured value of a quantity and its true value. Such variability can
arise through a whole plethora of mechanisms, which are often unique to
the study design and difficult to avoid\textsuperscript{27,28}. Array
technology is now commonly used to obtain high throughput phenotyping at
low cost, but they come with the problem of having imperfect resolution,
for instance methylation levels as measured by the Illumina450k chip are
prone to have some amount of noise around the true
value\textsuperscript{29,30}. Relatedly, if the measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{31}.

Measurement error will of course arise in more low-tech data too, for
example when measuring BMI one is typically interested in using this as
a proxy for adiposity, but it is clear that the correlation between BMI
and underlying adiposity is not perfect\textsuperscript{32}. A similar
problem of biological misspecification is unavoidable in disease
diagnosis, and measuring behaviour such as smoking or diet is
notoriously difficult to do accurately. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically incur both
changes in scale and imprecision (noise) compared to the original
variable. The sources of measurement error are not limited to this
list\textsuperscript{28}, and its impact has been explored in the
context of mediation analysis in the epidemiologyical literature
extensively\textsuperscript{6,21,24}.

Given the near-ubiquitous presence of measurement error in biological
data it is vital to understand its impact on the tools we use for causal
inference. Here we show using theory and simulations how
non-differential measurement error and measurement imprecision can lead
to unreliable causal inference in the mediation-based CIT method. We
also present an extension to MR that allows researchers to ascertain the
causal direction even when the biology of the instrument is not fully
understood. This improves the utility of MR in cases where mediation
based methods might have otherwise been used preferentially, and we
apply this new method to infer the direction of causation between DNA
methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{33} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{8} is implemented in the R package
\emph{R/cit}\textsuperscript{9}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
\end{aligned}
\]

The following tests are then performed:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \(H_0: cov(g, x) = 0; H_1: cov(g, x) \neq 0\); \emph{the SNP
  associates with the exposure}
\item
  \(H_0: cov(g, y) = 0; H_1: cov(g, y) \neq 0\); \emph{the SNP
  associates with the outcome}
\item
  \(H_0: cov(x, y) = 0; H_1: cov(x, y) \neq 0\); \emph{the exposure
  associates with the outcome}
\item
  \(H_0: cov(g, y - \hat{y}) \neq 0; cov(g, y - \hat{y}) = 0\);
  \emph{the SNP is independent of the outcome when the outcome is
  adjusted for the exposure}
\end{enumerate}

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for the \(x\), where \(x\) is assumed
to mediate the association between the SNP and the outcome. If all four
tests reject the null hypothesis then it is inferred that \(x\) is
causally related to \(y\). The CIT measures the strength of causality by
generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
(least extreme) p-value of the four tests, the intuition being that
causal inference is only as strong as the weakest link in the chain of
tests.

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the correct model of \(x\) causing \(y\), and for incorrect model of
\(y\) causing \(x\). For some significance threshold \(\alpha\), the
existence of causality and its direction was inferred based on the
following scenarios:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then the correct model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then the incorrect model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then an alternative causal model is accepted
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then no inference is made
\end{itemize}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we should stress that for real analyses
it is not good practice to use p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds.

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. Assuming the following model

\[
y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
\]

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{34}.

The direction of the causal association was inferred as follows.
Assuming no horizontal pleiotropy (i.e. \(g\) only has an effect on one
of the variables through the other variable) it is desirable to know
which of the variables, \(x\) or \(y\), it has a direct influence on.
This can be achieved by assessing which of the two variables has the
biggest absolute correlation with \(g\) (Appendix 2). This test can be
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{35}. It is calculated as

@@

and a p-value is generated from the \(Z\) value to indicate the
probability of obtaining the observed difference in correlations
\(\rho_{gx}\) and \(\rho_{gy}\) under the null hypothesis that both
correlations are identical. The existence of causality and its direction
is inferred based on the following scenarios:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
  a causal association for the correct model is accepted
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
  a causal association for the incorrect model is accepted
\item
  Otherwise no inference is made
\end{itemize}

For the purposes of compiling simulation results, we use an arbitrary
\(\alpha = 0.05\) value.

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{19} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (or from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

\[
Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
\]

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{36}.

@@

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

where \(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), \(\alpha_{mx}\) and
\(\beta_{mx}\) are parameters that represent non-differential
measurement error into the exposure variable \(x\), and \(\alpha_{my}\)
and \(\beta_{my}\) are parameters for non-differential measurement error
in the outcome \(y\).

All \(\alpha\) values were set to 0, and \(\beta\) values set to 1.
Normally distribted values of \(\epsilon_*\) were generated such that

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Simulations using each
of these sets of variables were performed 100 times, and the CIT and MR
methods were applied to each in order to evaluate the causal association
of the simulated variables. Similar patterns of results were obtained
for different values of \(cor(g, x)\).

\subsection{Two sample MR}\label{two-sample-mr}

Two sample MR\textsuperscript{19} was performed using the summary
statistics presented in\textsuperscript{18} for each of 458 associations
between DNA methylation and gene expression levels. Because only summary
statistics were available (effect, standard error, effect allele, sample
size, p-values) for the instrumental SNP on the methylation and gene
expression levels, the Steiger test of two independent correlations was
used to infer the direction of causality for each of the associations.
The Wald ratio test was then used to estimate the causal effect size for
the estimated direction for each association.

@@

\subsubsection{The influence of measurement error on
mediation}\label{the-influence-of-measurement-error-on-mediation}

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_O = f(x)\). For
example, we can define
\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\), and \(\epsilon_{mx}\) represents the imprecision
in the measurement of \(x\). Here the true value of the exposure is
partially explained by the genetic instrument, \(g\), such that

@@

where \(\beta_g\) is the effect of the SNP on the exposure, and
\(\epsilon_g\) is the normally distributed residual value of \(x\); and
the outcome is partially explained by the exposure

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement
error is introduced it can be shown using basic covariance properties
(Appendix 1) that
